---
title: "Streaked Horned Lark detectability and abundance: preliminary results"
author: "John D. Lloyd"
format: html
editor: visual
bibliography: references.bib
---

```{r}
#| echo: false
#| warning: false
#| output: false
library(tidyverse)
library(readxl)
library(lubridate)
library(suncalc)
library(ggpmisc)
library(unmarked)
library(AICcmodavg)
library(ubms)
library(knitr)
# Review and organize data, changing formats and variable names as needed.
#surveyData <- read_xlsx("~/Documents/GitHub/hornedLarks/WV_SurveyOutput.xlsx")
surveyData <- read_csv("~/Documents/GitHub/hornedLarks/WV_SHLA_22.csv")
names(surveyData)[1] <- 'surveyEvent'
#surveyData$Count_Date <- mdy(surveyData$Count_Date)
surveyData$Count_Date <- mdy(surveyData$Survey_Date)
#surveyData$Start_Time <- hms(surveyData$Start_Time)
surveyData$Start_Time <- hms(surveyData$Survey_Time)
surveyData$Site_ID <- factor(surveyData$Site_ID)
surveyData$Observer <- factor(surveyData$Observer)
surveyData$Sky_Code <- factor(surveyData$Sky_Code, levels = c("0","1","2","3","4"), 
                              labels = c("Clear","Partly cloudy","Mostly cloudy","Fog or smoke","Drizzle"))
surveyData$Sex <- factor(surveyData$Sex, levels = c("M","F","U"), labels = c("Male","Female","Unknown"))
surveyData$Age <- factor(surveyData$Age, levels = c("A", "J"), labels = c("Adult", "Juvenile"))
surveyData$`Distance Band` <- factor(surveyData$`Distance Band`)
names(surveyData)[19] <- 'distanceBand'
surveyData$Interval_1 <- ifelse(surveyData$Interval_1 == "X", NA, surveyData$Interval_1)
surveyData$Interval_2 <- ifelse(surveyData$Interval_2 == "X", NA, surveyData$Interval_2)
surveyData$Interval_3 <- ifelse(surveyData$Interval_3 == "X", NA, surveyData$Interval_3)
surveyData$Interval_4 <- ifelse(surveyData$Interval_4 == "X", NA, surveyData$Interval_4)

surveyData$Interval_1 <- factor(surveyData$Interval_1, levels = c("C","S","V"), labels = 
                                  c("Calling", "Singing", "Visual"))
surveyData$Interval_2 <- factor(surveyData$Interval_2, levels = c("C","S","V"), labels = 
                                  c("Calling", "Singing", "Visual"))
surveyData$Interval_3 <- factor(surveyData$Interval_3, levels = c("C","S","V"), labels = 
                                  c("Calling", "Singing", "Visual"))
surveyData$Interval_4 <- factor(surveyData$Interval_4, levels = c("C","S","V"), labels = 
                                  c("Calling", "Singing", "Visual"))

surveyData$dayOfYear <- yday(surveyData$Count_Date) # create a day-of-year variable for analysis

## Add a "first detected by..." column to survey data:
surveyData <-
  surveyData %>%
  filter(!is.na(Sex)) %>%
  select(Sex, Lark_ID, Interval_1, Interval_2, Interval_3, Interval_4) %>%
  pivot_longer(., cols = 3:6, names_to = "interval", values_to = "detection") %>%
  group_by(Lark_ID, Sex) %>%
  filter(!is.na(detection)) %>%
  summarise(firstDet = first(detection)) %>%
  right_join(., surveyData, by = 'Lark_ID', keep = F) %>%
  select(!(Sex.x)) %>%
  rename(Sex = Sex.y)

## Add a "first detected interval" column
surveyData <-
surveyData %>%
  filter(!is.na(Sex)) %>%
  select(Sex, Lark_ID, Interval_1, Interval_2, Interval_3, Interval_4) %>%
  pivot_longer(., cols = 3:6, names_to = "interval", values_to = "detection") %>%
  group_by(Lark_ID, Sex) %>%
  filter(!is.na(detection)) %>%
  summarise(firstDet = first(interval)) %>%
  mutate(firstInterval = ifelse(firstDet == "Interval_1",1,
                                ifelse(firstDet == "Interval_2", 2,
                                       ifelse(firstDet == "Interval_3", 3,
                                              ifelse(firstDet == "Interval_4", 4, NA))))) %>%
  right_join(., surveyData, by = 'Lark_ID', keep = F) %>%
  select(!(Sex.x)) %>%
  select(!firstDet.x) %>%
  rename(firstDet = firstDet.y, Sex = Sex.y)

# Get sunrise times to look at effect of survey timing in detections,
# here using the Corvallis airport as the location. We could calculate
# sunrise based on the lat/long of each survey point, but they are 
# all going to have roughly the same sunrise so for ease I'm simply estimating
# sunrise at a single, central location.

# calculate sunrise times at Corvallis Airport on each survey date.
sunriseTimes <- getSunlightTimes(date = surveyData$Count_Date, lat = 44.50, lon = -123.28,
                                 keep = c("sunrise"), tz="America/Los_Angeles")

#create a new variable in surveyData that has sunrise matched to the survey point.
#careful, here, because there is no matching function (i.e., this only works if the
#two data frames are sorted in the same order. This should be the case unless you
#sort one after calling this function).
surveyData$sunrise <- sunriseTimes$sunrise
#subtract the two times to get decimal hours after sunrise.
surveyData$mas <- 60*((surveyData$Start_Time@hour + surveyData$Start_Time@minute/60) -
                        (hour(surveyData$sunrise) + minute(surveyData$sunrise)/60))
#clean up
rm(sunriseTimes)

# Read in survey location and habitat data
habitatData <- read_csv("tbl_survey_locations_exported_09_26_2022_pct_suitable_2021.csv")
habitatData$Site_ID <- factor(habitatData$Site_ID)

habitatAndSurveyData <-
habitatData %>%
  select(Site_ID,ln_UTM1083,lt_UTM1083,ln_WGS84,lt_WGS84, pct_suitable_2021) %>%
  right_join(.,surveyData, by = 'Site_ID', keep = F)

habitatAndSurveyData$Site_ID <- factor(habitatAndSurveyData$Site_ID)
habitatAndSurveyData$Observer <- factor(habitatAndSurveyData$Observer)

# Make encounter histories for detected birds across intervals
surveyData$encounterHistory <- paste(if_else(surveyData$Interval_1 == "None", 0,1),
                                     if_else(surveyData$Interval_2 == "None", 0,1),
                                     if_else(surveyData$Interval_3 == "None", 0,1),
                                     if_else(surveyData$Interval_4 == "None", 0,1),
                                     sep = "")


```

# Encounter rates

Streaked Horned Larks were detected at 13% (n = 28) of the 214 points surveyed in the Willamette Valley of Oregon in 2022 (@tbl-encounterRates).

```{r}
#| echo: false
#| warning: false
#| label: tbl-encounterRates
#| tbl-cap: Number of Streaked Horned Larks detected during surveys in the Willamette Valley of Oregon, 2022.
tableEncounterRates <-
surveyData %>%
  group_by(Site_ID) %>%
  summarise(larksDetected = first(Number_Detected)) %>%
  group_by(larksDetected) %>%
  summarise(count = n()) %>%
  mutate(freq = count/sum(count))

kable(as.data.frame(tableEncounterRates), digits = 2, col.names = c("No. of larks detected", "No. of survey points", "Frequency"), caption = "Number of Streaked Horned Larks detected during surveys in the Willamette Valley of Oregon, 2022.")
```

However, encounter rates of singing males were slightly lower (@tbl-encounterRatesSingingMales). This is significant because it determines the sample size for the distance-sampling analysis, which is generally conducted only on singing males.

```{r}
#| echo: false
#| warning: false
#| label: tbl-encounterRatesSingingMales
#| tbl-cap: Number of singing Streaked Horned Larks detected during surveys in the Willamette Valley of Oregon, 2022.
tableEncounterRatesSingingMales <-
surveyData %>%
  mutate(singMale = ifelse(firstDet == "Singing", 1, 0)) %>%
  mutate(singMale = ifelse(is.na(singMale), 0, singMale)) %>%
group_by(Site_ID) %>%
  summarise(larksDetected = sum(singMale)) %>%
  group_by(larksDetected) %>%
  summarise(count = n()) %>%
  mutate(freq = count/sum(count))

kable(as.data.frame(tableEncounterRatesSingingMales), digits = 2, col.names = c("No. of singing larks detected", "No. of survey points", "Frequency"), caption = "Number of singing Streaked Horned Larks detected during surveys in the Willamette Valley of Oregon, 2022.")
```

# Estimating perceptibility of singing larks with distance sampling

Distance sampling estimates the extent of the decline in an observer's ability to detect a bird as a function of the distance between the observer and bird. This parameter, referred to here as perceptibility, estimates the probability of detecting a bird within the survey area given that it was present during the survey and that it made itself available for detection, for example by singing. It does not account for birds present but unavailable for detection, for example because they did not sing and remained out of view.

Because distance sampling is sensitive to individual heterogeneity in the shape of the detection function, I adopted standard practice and limited this analysis to singing males. Non-singing males and females may have a different detection function (@tbl-sexAndCue) and so including them in the analysis could introduce substantial bias. This decision reduces the sample size considerably; with a greater number of detections, models that include sex and detection type as covariates could be fit.

```{r}
#| echo: false
#| warning: false
#| label: tbl-sexAndCue
#| tbl-cap: Streaked Horned Lark detections by sex and cue during surveys in the Willamette Valley of Oregon, 2022.
tableSexAndCue <- surveyData %>%
  filter(!is.na(Sex)) %>%
  select(Sex, Lark_ID, Interval_1, Interval_2, Interval_3, Interval_4) %>%
  pivot_longer(., cols = 3:6, names_to = "interval", values_to = "detection") %>%
  group_by(Lark_ID, Sex) %>%
  filter(!is.na(detection)) %>%
  summarise(firstDet = first(detection)) %>%
  group_by(Sex, firstDet) %>%
  summarise(numDetected = n())

kable(as.data.frame(tableSexAndCue), col.names = c("Sex", "Detection type", "Number"), caption = "Streaked Horned Lark detections by sex and cue during surveys in the Willamette Valley of Oregon, 2022.")

```

```{r}
#| echo: false
#| warning: false
#| output: false
## Create a new variable called 'distance', which translates the distance_band information into the midpoint of the distance.
## Then remove all other variables.
  ## THIS EXCLUDES ALL NON-SINGING MALES ##
dists <-
  surveyData %>%
  group_by(Site_ID) %>%
  mutate(distance = ifelse(distanceBand == 1, 12.5,
                           ifelse(distanceBand == 2, 61,
                                  ifelse(distanceBand == 3, 150,
                                         ifelse(distanceBand == 4, 300, NA))))) %>%
  select(Site_ID, distance, Sex, firstDet) %>%
  filter(!is.na(distance), firstDet == "Singing", Sex == "Male")

## Note here that we need the "as.data.frame" argument because 'dists' is a tidyverse tibble, 
## and unmarked doesn't seem to like tibbles. This forces it into a standard R data frame.
yDat <- formatDistData(distData = as.data.frame(dists), distCol = "distance", transectNameCol = "Site_ID", 
                       dist.breaks = c(0,25,100,200,400))

## Create a data frame of site-level covariates.
covs <-
  surveyData %>%
  group_by(Site_ID) %>%
  summarise(site = first(Site_ID),
            observer = first(Observer),
            temp = first(Temp),
            avgNoise = first(Avg_Noise),
            dayOfYear = first (dayOfYear),
            mas = first(mas))


umf <- unmarkedFrameDS(y = as.matrix(yDat), siteCovs = as.data.frame(covs),
                       survey = "point", dist.breaks = c(0,25,100,200,400), unitsIn = "m")
```

Most of the 31 singing males encountered during surveys were detected in the second and third distance bands (@fig-detectionDistances). That only a single detection was registered in the first distance band is not unexpected given that the area included in this band is substantially smaller than the area included in the other distance bands.

```{r}
#| echo: false
#| warning: false
#| label: fig-detectionDistances
#| fig-cap: "Streaked Horned Lark detection distances during surveys in the Willamette Valley of Oregon, 2022."
#| fig-cap-location: bottom
hist(umf, freq = TRUE, xlab = "Distance (m)", main = NULL, cex.lab = 0.8, cex.axis = 0.8)

```

```{r}
#| output: false
#| warning: false
#| echo: false
# Fitting models.
# Half-normal, null
hnNull <- distsamp(~1~1, umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

#Half-normal, MAS
hnMAS <- distsamp(~mas ~1, umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

# Half-normal, Day of year
hnDay <- distsamp(~dayOfYear ~1, data = umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

# Half-normal, noise
hnNoise <- distsamp(~avgNoise ~1, data = umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

# Half-normal, Temp
hnTemp <- distsamp(~temp ~1, data = umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

# Hazard-rate models
haNull <- distsamp(~1 ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")
haNoise <- distsamp(~avgNoise ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")
haTemp <- distsamp(~temp ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")
haDay <- distsamp(~dayOfYear ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")
haMAS <- distsamp(~mas ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")

fmList <- list("haNull" = haNull, "haDay" = haDay, "haNoise" = haNoise, "haMAS" = haMAS, "haTemp" = haTemp,
               "hnNull" = hnNull, "hnDay" = hnDay, "hnNoise" = hnNoise, "hnMAS" = hnMAS, "hnTemp" = hnTemp)
```

I considered models in the distance-sampling analysis that included individual effects of four potential covariates on perceptibility: day of the year ("day"), ambient noise ("noise"), temperature at the time of the survey ("temp"), and time of day (quantified as the number of minutes after sunrise ("MAS") that the survey began). I also considered null models in which no covariates on perceptibility were included. I did not include any covariates on abundance, both because of sample size concerns and because we have yet to generate any covariates that might predict variation in abundance. I did not include models with more than one covariate due to the small sample size. I considered two suites of these individual-covariate models, one that assumed a half-normal ("hn") detection function and one that assumed a hazard-rate ("ha") detection function.

The best-supported model in the distance-sampling analysis included an effect of day-of-the-year and a half-normal detection function (@tbl-distanceAIC). Model fit was adequate (all P-values for goodness-of-fit testing \> 0.12). None of the other covariates had a strong influence on perceptibility. In general, the hazard-rate models performed no better than the half-normal models, but required estimating one additional parameter and so tended to rank beneath half-normal models by about 2 AIC values.

```{r}
#| warning: false
#| echo: false
#| label: tbl-distanceAIC
#| tbl-cap: Model selection results for distance-sampling analysis of Streaked Horned Lark detections during surveys in the Willamette Valley of Oregon, 2022.
tableDistanceAIC <- aictab(cand.set = fmList, second.ord = T, sort = T)
kable(as.data.frame(tableDistanceAIC), digits = 2, col.names = c("Model","No. parameters", "AICc", "Delta AICc", "Model likelihood", "AICc weight", "Log-likelihood", "Cum. weight"), caption = "Model selection results for distance-sampling analysis of Streaked Horned Lark detections during surveys in the Willamette Valley of Oregon, 2022.")
```

```{r}
#| warning: false
#| echo: false
#| output: false
## Predict seasonal changes in P-hat
## Loop function for calculating P-hat across days of the season
distanceCovPred <- matrix(nrow = 24, ncol = 7) #create a matrix for output
colnames(distanceCovPred) <- c("sigma", "ea", "er", "p","lowerCI", "upperCI", "Day") #give the columns names
for (i in 1:24) { #this loops through 24 days of the year to generate estimates of p-hat
  distanceCovPred[[i,1]] <- backTransform(linearComb(hnDay['det'], c(1, i+157)))@estimate # this is a kludge to index days (i.e., lowest value = day 158)
  distanceCovPred[[i,2]] <- 2*pi * integrate(grhn, 0, 400, sigma=distanceCovPred[[i,1]])$value # effective area
  distanceCovPred[[i,3]] <- sqrt(distanceCovPred[[i,2]] / pi) # effective radius
  distanceCovPred[[i,4]] <- distanceCovPred[[i,2]] / (pi*400^2) #detection probability
  getPcovsL <- function(hnDay) { #this defines the function that 'parboot' will use
    sig <- backTransform(linearComb(hnDay['det'], c(1, i+157)))@estimate
    ea <- 2*pi * integrate(grhn, 0, 400, sigma=sig)$value # effective area
    er <- sqrt(ea / pi) # effective radius
    p <- ea / (pi*400^2) #detection probability
    out <- c(p = p , er = er)
  } #following lines are for pulling out upper and lower 95% Ci from bootstrap; parallel process didn't work, so had to set cores = 1
  distanceCovPred[[i,5]] <- quantile(parboot(hnDay,getPcovsL, nsim = 100, report = 100, ncores = 1)@t.star[,1],probs = 0.025)
  distanceCovPred[[i,6]] <- quantile(parboot(hnDay,getPcovsL, nsim = 100, report = 100, ncores = 1)@t.star[,1], probs = 0.975)
  }

distanceCovPred[,7]<- seq(158, 181, 1)


```

The effect of day-of-year on perceptibility is negative, decreasing substantially throughout the season, such that the probability of detecting a lark, given that it sings, approaches zero by late June (@fig-dayOfYearEffect). The predicted perceptibility on the median day of surveys was 0.06 (95% CI = 0.03 - 0.10).

```{r}
#| warning: false
#| echo: false
#| label: fig-dayOfYearEffect
#| fig-cap: "Streaked Horned Lark percebtibility declined over the course of the breeding season during surveys in the Willamette Valley of Oregon, 2022. Solid line is the predicted value of perceptibility and gray shaded area is the 95% confidence interval."
#| fig-cap-location: bottom
ggplot(data = as.data.frame(distanceCovPred), aes(x = Day, y = p)) +
  geom_ribbon(aes(ymin = lowerCI, ymax = upperCI), fill = "grey80", alpha = 0.5) + 
  geom_line(aes(x = Day, y = p), color = "black") + 
  xlab("Day of year") + ylab ("Probability of detection") + 
  theme_bw()
```

With the low estimated perceptibility of larks, the resulting abundance estimate from the best distance model was also quite high: 4.3 larks per km^2^ (95% CI = 2.2 - 6.9).

# Estimating availability of singing larks with removal models

Availabiity is another component of detectability. Availability refers to the probability that a bird present in the count area during the survey produces a cue that can be detected by the observer. In this analysis, I estimated availability in two ways. First, by using a removal model, in which each individual detected is considered to have been removed from the population of individuals present once it produces a cue that is detected by the observer. This approach does not retain the full encounter history generated by the time-of-detection survey method. As a supplement, therefore, I also estimated the probability of availability using the full encounter history for each detected individual (i.e., the series of zeros and ones that indicate whether an individual was observed during each of the 4 survey intervals). In both cases, I considered models with covariates for availability only; I did not consider covariates on abundance.

```{r}
#| output: false
#| warning: false
#| echo: false


## Incorporating removal models
encounters <-
  surveyData %>%
  group_by(Site_ID) %>%
  mutate(interval1 = ifelse(firstInterval == 1 & Sex == "Male" & firstDet == "Singing", 1, 0),
         interval2 = ifelse(firstInterval == 2 & Sex == "Male" & firstDet == "Singing", 1, 0),
         interval3 = ifelse(firstInterval == 3 & Sex == "Male" & firstDet == "Singing", 1, 0),
         interval4 = ifelse(firstInterval == 4 & Sex == "Male" & firstDet == "Singing", 1, 0)) %>%
  select(Site_ID, interval1, interval2, interval3, interval4, Sex, firstDet) %>%
  group_by(Site_ID) %>%
  summarise(interval1 = sum(interval1), # this model wants summed # of birds per interval
            interval2 = sum(interval2), # and has to match the distance data
            interval3 = sum(interval3),
            interval4 = sum(interval4)) 

yRemoval <- matrix(nrow = 214, ncol = 4)
rownames(yRemoval) <- encounters$Site_ID
yRemoval <- cbind(encounters[,2:5])
yRemoval[is.na(yRemoval)] <- 0
## Create a new variable called 'distance', which translates the distance_band information into the midpoint of the distance.
## Then remove all other variables.
## THIS INCLUDES ONLY SINGING MALES ##
distsRemoval <-
  surveyData %>%
  group_by(Site_ID) %>%
  mutate(distance = ifelse(distanceBand == 1, 12.5,
                           ifelse(distanceBand == 2, 61,
                                  ifelse(distanceBand == 3, 150,
                                         ifelse(distanceBand == 4, 300, NA))))) %>%
  select(Site_ID, distance, Sex, firstDet) %>%
  filter(!is.na(distance), firstDet == "Singing", Sex == "Male")

## Note here that we need the "as.data.frame" argument because 'dists' is a tidyverse tibble, 
## and unmarked doesn't seem to like tibbles. This forces it into a standard R data frame.
yDatDistance <- formatDistData(distData = as.data.frame(distsRemoval), distCol = "distance", transectNameCol = "Site_ID", 
                       dist.breaks = c(0,25,100,200,400))



umfDR <- unmarkedFrameGDR(yDistance = as.matrix(yDatDistance), yRemoval = as.matrix(yRemoval), numPrimary = 1,
                          siteCovs = covs, dist.breaks = c(0,25,100,200,400), unitsIn = "m")

removalFrame <- unmarkedFrameMPois(y = yRemoval, siteCovs = covs, type = "removal")
removalNull <- multinomPois(~1 ~1, data = removalFrame)
removalDay <- multinomPois(~scale(dayOfYear) ~1, data = removalFrame)
removalTemp <- multinomPois(~temp ~1, data = removalFrame)
removalNoise <- multinomPois(~avgNoise ~1, data = removalFrame)
removalMAS <- multinomPois (~mas ~1, data = removalFrame)

fmRemovalList <- list("removalNull" = removalNull, "removalDay" = removalDay, "removalTemp" = removalTemp,
               "removalNoise" = removalNoise, "removalMAS" = removalMAS)
```

The best-supported model of availability from the removal analysis indicated an effect of day-of-the-year on availability (@tbl-removalAIC).

```{r}
#| warning: false
#| echo: false
#| label: tbl-removalAIC
#| tbl-cap: Model selection results for removal analysis of Streaked Horned Lark detections during surveys in the Willamette Valley of Oregon, 2022.
fmRemovalList <- list("removalNull" = removalNull, "removalDay" = removalDay, "removalTemp" = removalTemp,
               "removalNoise" = removalNoise, "removalMAS" = removalMAS)
tableRemovalAIC <- aictab(cand.set = fmRemovalList, second.ord = T, sort = T)
kable(as.data.frame(tableRemovalAIC), digits = 2, col.names = c("Model","No. parameters", "AICc", "Delta AICc", "Model likelihood", "AICc weight", "Log-likelihood", "Cum. weight"), caption = "Model selection results for removal analysis of Streaked Horned Lark detections during surveys in the Willamette Valley of Oregon, 2022.")
```

Availability for detection declined over the course of the season (@fig-dayOfYearEffectRemoval). The estimated probability of availability on the median day of surveys was 0.14 (SE = 0.06). Estimated abundance from this model was substantially lower than than produced by distance sampling, in part due to the higher estimate of detectability: 0.6 birds per km^2^ (SE = 0.09).

```{r}
#| warning: false
#| echo: false
#| label: fig-dayOfYearEffectRemoval
#| fig-cap: "Removal models indicated that Streaked Horned Lark availability declined over the course of the breeding season during surveys in the Willamette Valley of Oregon, 2022. Solid line is the predicted value of perceptibility and gray shaded area is the 95% confidence interval."
#| fig-cap-location: bottom
removalPredictPdata <- data.frame(dayOfYear = seq(min(covs$dayOfYear), max(covs$dayOfYear), by = 1))
removalPredictP <- predict(removalDay, type = "det", newdata = removalPredictPdata, appendData = TRUE) 
ggplot(data = removalPredictP, aes(x = dayOfYear, y = Predicted)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey80", alpha = 0.5) + 
  geom_line(aes(x = dayOfYear, y = Predicted), color = "black") + 
  xlab("Day of year") + ylab ("Availability for detection") + 
  theme_bw()

```

```{r}
#| output: false
#| warning: false
#| echo: false
## How do removal models compare to full encounter histories? Following code will treat the 
## surveys as capture/recapture efforts, thus retaining the full encounter history. This
## differs from the removal models, which truncate the history after the first detection.
captHistory <-
  encounters %>%
  replace(is.na(.), 0) %>%
  mutate(captureHistory = paste(interval1, interval2, interval3, interval4, sep = "")) 
  
captHistory$captureHistory <- factor(captHistory$captureHistory, 
                                     levels = c("0001","0010","0100","1000",
                                                "0011","0101","0110","1010","1001","1100",
                                                "0111","1011","1101","1110","1111"))
capSummary <- table(captHistory$Site_ID, captHistory$captureHistory)

crPiFun <- function(p) {
  p1 <- p[,1]
  p2 <- p[,2]
  p3 <- p[,3]
  p4 <- p[,4]
  cbind("0001" = (1-p1) * (1-p2) * (1-p3) * p4,
        "0010" = (1-p1) * (1-p2) * p3 * (1-p4),
        "0100" = (1-p1) * p2 * (1-p3) * (1-p4),
        "1000" = p1 * (1-p2) * (1-p3) * (1-p4),
        "0011" = (1-p1) * (1-p2) * p3 * p4,
        "0101" = (1-p1) * p2 * (1-p3) * p4,
        "0110" = (1-p1) * p2 * p3 * (1-p4),
        "1010" = p1 * (1-p2) * p3 * (1-p4),
        "1001" = p1 * (1-p2) * (1-p3) * p4,
        "1100" = p1 * p2 * (1-p3) * (1-p4),
        "0111" = (1-p1) * p2 * p3 * p4,
        "1011" = p1 * (1-p2) * p3 * p4,
        "1101" = p1 * p2 * (1-p3) * p4,
        "1110" = p1 * p2 * p3 * (1-p4),
        "1111" = p1 * p2 * p3 * p4)
}

o2y <- matrix(1, 4, 15)

class(capSummary) <- "matrix"
umf.cr1 <- unmarkedFrameMPois(y = capSummary, siteCovs = covs, piFun = "crPiFun", obsToY = o2y)
crNull <- multinomPois(~1 ~1, umf.cr1, engine = "R")
crDay <- multinomPois(~dayOfYear ~1, umf.cr1, engine = "R")
crTemp <- multinomPois(~temp ~1, data = umf.cr1, engine = "R")
crNoise <- multinomPois(~avgNoise ~1, data = umf.cr1, engine = "R")
crMAS <- multinomPois (~mas ~1, data = umf.cr1, engine = "R")

fmCRList <- list("crNull" = crNull, "crDay" = crDay, "crTemp" = crTemp,
                      "crNoise" = crNoise, "crMAS" = crMAS)
```

Analyzing the data in mark-recapture format - that is, retaining the full encounter history for each individual - yielded similar results. The best-supported model again indicated an effect of day-of-the-year on availability (@tbl-crAIC).

```{r}
#| warning: false
#| echo: false
#| label: tbl-crAIC
#| tbl-cap: Model selection results for mark-recapture analysis of Streaked Horned Lark detections during surveys in the Willamette Valley of Oregon, 2022.
tableCRAIC <- aictab(cand.set = fmCRList, second.ord = T, sort = T)
kable(as.data.frame(tableCRAIC), digits = 2, col.names = c("Model","No. parameters", "AICc", "Delta AICc", "Model likelihood", "AICc weight", "Log-likelihood", "Cum. weight"), caption = "Model selection results for removal analysis of Streaked Horned Lark detections during surveys in the Willamette Valley of Oregon, 2022.")
```

Availability for detection declined over the course of the season (@fig-dayOfYearEffectCR). The estimated probability of availability on the median day of surveys was 0.13 (SE = 0.05). Estimated abundance was 0.35 birds per km^2^ (SE = 0.06). Thus, both removal and mark-recapture models yield similar estimates of availability and abundance, with a slight gain in precision associated with use of the full encounter history.

```{r}
#| warning: false
#| echo: false
#| label: fig-dayOfYearEffectCR
#| fig-cap: "Mark-recapture analyses indicated that Streaked Horned Lark availability declined over the course of the breeding season during surveys in the Willamette Valley of Oregon, 2022. Solid line is the predicted value of perceptibility and gray shaded area is the 95% confidence interval."
#| fig-cap-location: bottom
crPredictPdata <- data.frame(dayOfYear = seq(min(covs$dayOfYear), max(covs$dayOfYear), by = 1))
crPredictP <- predict(crDay, type = "det", newdata = crPredictPdata, appendData = TRUE) 
ggplot(data = crPredictP, aes(x = dayOfYear, y = Predicted)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey80", alpha = 0.5) + 
  geom_line(aes(x = dayOfYear, y = Predicted), color = "black") + 
  xlab("Day of year") + ylab ("Availability for detection") + 
  theme_bw()

```

# Inference about overall detectability from distance sampling, removal models, and mark-recapture models.

Assuming independence in the probabilities of availability and perceptibility, their product provides an ad-hoc estimate of overall average detectability on the median day of surveys: 0.14 (probability of availability) \* 0.06 (probability a lark is detected given that it sings) = 0.01. Because of the seasonal changes in both availability and perceptibility, detectability will be higher than this earlier in the year but lower later in the year.

# Joint distance-sampling and removal models

The hierarchical model of Amundson et al. [-@amundson2014] allows for a unified approach to estimating detectability via distance sampling and time-removal models. I used a similar exploratory approach to constructing the model set, using day-of-year as a distance covariate given its strong support in the distance-sampling analysis. I then added each of the potential covariates to the time-removal portion of the model. For comparison, I also analyzed a null model and two models that included no covariates for the distance model but used time-of-day and time-of-year as covariates on the time-removal portion of the model. The best-supported model included an effect of day-of-year on the distance model and no covariates on the removal portion of the model. None of the covariates had especially strong effects on model fit, as the null model was only \~ 2 AIC points below the best-supported model.

[Table 4: Model selection results for hierachical distance-sampling and time-removal models of Streaked Horned Lark detections during surveys in the Willamette Valley of Oregon, 2022.]{style="text-align: center"}

| Distance covariate | Removal covariate | AICc   | ðž“AICc |
|:------------------:|:-----------------:|--------|-------|
|    Day-of-year     |        \-         | 296.45 | 0     |
|    Day-of-year     |    Time-of-day    | 297.98 | 1.53  |
|    Day-of-year     |    Day-of-year    | 298.04 | 1.59  |
|    Day-of-year     |       Noise       | 298.44 | 1.99  |
|    Day-of-year     |    Temperature    | 298.44 | 1.99  |
|         \-         |        \-         | 298.66 | 2.21  |
|         \-         |    Day-of-year    | 298.72 | 2.27  |
|         \-         |    Time-of-day    | 300.09 | 3.64  |

From the best model, the estimated probability of detecting a lark given that it sang was 0.06, the estimated probability of a lark being available given that it was present during the survey was 0.22, and the overall probability of detection was 0.01. This model yielded an abundance estimate of 37.7 larks per km^2^.

# Discussion

Analysis of the 2022 data yielded very low estimates of detectability, which in turn leads to unreasonably high and imprecise estimates of abundance (raw counts are divided by the estimated detectability to yield an estimate of abundance; all things being equal, apparent abundance will increase as detectability declines). Model fit was adequate, however, which suggests that small sample size may be the primary driver of the apparently low detection rates. An important goal for subsequent years should be to increase the encounter rate, as only 10% of points surveyed yielded a singing male.

Increasing the encounter rate could be accomplished via two approaches. First, these preliminary findings suggest that beginning and completing surveys earlier in the year may be beneficial because of the apparently steep decline in availability of larks later in the season. Surveying when singing is more frequent might increase encounter rates by avoiding false absences that arise when surveys yield no detections even though birds were present. Second, refining the definition of the sampling frame may help reduce the number of unoccupied points that are surveyed, although this assumes that points without lark detections in 2022 were unsuitable, rather than being suitable but unoccupied due to small population size.

The time-removal and capture-recapture approaches to estimating availability and abundance performed equally well, with the time-removal models appearing somewhat less precise. However, the time-removal models are less demanding on field observers, and it might be worthwhile to consider whether the burden of tracking individual larks across time intervals is such that field methods should be simplified.
