---
title: "Streaked Horned Lark Preliminary Analysis"
author: "John D Lloyd"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Data import and formatting, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
# This chunk of code reads and formats the original data files.
library(tidyverse)
library(readxl)
library(lubridate)
library(suncalc)
library(ggpmisc)
library(unmarked)
library(AICcmodavg)

# Review and organize data, changing formats and variable names as needed.
#surveyData <- read_xlsx("~/Documents/GitHub/hornedLarks/WV_SurveyOutput.xlsx")
surveyData <- read_csv("~/Documents/GitHub/hornedLarks/WV_SHLA_22.csv")
names(surveyData)[1] <- 'surveyEvent'
#surveyData$Count_Date <- mdy(surveyData$Count_Date)
surveyData$Count_Date <- mdy(surveyData$Survey_Date)
#surveyData$Start_Time <- hms(surveyData$Start_Time)
surveyData$Start_Time <- hms(surveyData$Survey_Time)
surveyData$Site_ID <- factor(surveyData$Site_ID)
surveyData$Observer <- factor(surveyData$Observer)
surveyData$Sky_Code <- factor(surveyData$Sky_Code, levels = c("0","1","2","3","4"), 
                              labels = c("Clear","Partly cloudy","Mostly cloudy","Fog or smoke","Drizzle"))
surveyData$Sex <- factor(surveyData$Sex, levels = c("M","F","U"), labels = c("Male","Female","Unknown"))
surveyData$Age <- factor(surveyData$Age, levels = c("A", "J"), labels = c("Adult", "Juvenile"))
surveyData$`Distance Band` <- factor(surveyData$`Distance Band`)
names(surveyData)[19] <- 'distanceBand'
surveyData$Interval_1 <- ifelse(surveyData$Interval_1 == "X", NA, surveyData$Interval_1)
surveyData$Interval_2 <- ifelse(surveyData$Interval_2 == "X", NA, surveyData$Interval_2)
surveyData$Interval_3 <- ifelse(surveyData$Interval_3 == "X", NA, surveyData$Interval_3)
surveyData$Interval_4 <- ifelse(surveyData$Interval_4 == "X", NA, surveyData$Interval_4)

surveyData$Interval_1 <- factor(surveyData$Interval_1, levels = c("C","S","V"), labels = 
                                  c("Calling", "Singing", "Visual"))
surveyData$Interval_2 <- factor(surveyData$Interval_2, levels = c("C","S","V"), labels = 
                                  c("Calling", "Singing", "Visual"))
surveyData$Interval_3 <- factor(surveyData$Interval_3, levels = c("C","S","V"), labels = 
                                  c("Calling", "Singing", "Visual"))
surveyData$Interval_4 <- factor(surveyData$Interval_4, levels = c("C","S","V"), labels = 
                                  c("Calling", "Singing", "Visual"))

surveyData$dayOfYear <- yday(surveyData$Count_Date) # create a day-of-year variable for analysis

## Add a "first detected by..." column to survey data:
surveyData <-
  surveyData %>%
  filter(!is.na(Sex)) %>%
  select(Sex, Lark_ID, Interval_1, Interval_2, Interval_3, Interval_4) %>%
  pivot_longer(., cols = 3:6, names_to = "interval", values_to = "detection") %>%
  group_by(Lark_ID, Sex) %>%
  filter(!is.na(detection)) %>%
  summarise(firstDet = first(detection)) %>%
  right_join(., surveyData, by = 'Lark_ID', keep = F) %>%
  select(!(Sex.x)) %>%
  rename(Sex = Sex.y)

## Add a "first detected interval" column
surveyData <-
surveyData %>%
  filter(!is.na(Sex)) %>%
  select(Sex, Lark_ID, Interval_1, Interval_2, Interval_3, Interval_4) %>%
  pivot_longer(., cols = 3:6, names_to = "interval", values_to = "detection") %>%
  group_by(Lark_ID, Sex) %>%
  filter(!is.na(detection)) %>%
  summarise(firstDet = first(interval)) %>%
  mutate(firstInterval = ifelse(firstDet == "Interval_1",1,
                                ifelse(firstDet == "Interval_2", 2,
                                       ifelse(firstDet == "Interval_3", 3,
                                              ifelse(firstDet == "Interval_4", 4, NA))))) %>%
  right_join(., surveyData, by = 'Lark_ID', keep = F) %>%
  select(!(Sex.x)) %>%
  select(!firstDet.x) %>%
  rename(firstDet = firstDet.y, Sex = Sex.y)

# Get sunrise times to look at effect of survey timing in detections,
# here using the Corvallis airport as the location. We could calculate
# sunrise based on the lat/long of each survey point, but they are 
# all going to have roughly the same sunrise so for ease I'm simply estimating
# sunrise at a single, central location.

# calculate sunrise times at Corvallis Airport on each survey date.
sunriseTimes <- getSunlightTimes(date = surveyData$Count_Date, lat = 44.50, lon = -123.28,
                                 keep = c("sunrise"), tz="America/Los_Angeles")

#create a new variable in surveyData that has sunrise matched to the survey point.
#careful, here, because there is no matching function (i.e., this only works if the
#two data frames are sorted in the same order. This should be the case unless you
#sort one after calling this function).
surveyData$sunrise <- sunriseTimes$sunrise
#subtract the two times to get decimal hours after sunrise.
surveyData$mas <- 60*((surveyData$Start_Time@hour + surveyData$Start_Time@minute/60) -
                        (hour(surveyData$sunrise) + minute(surveyData$sunrise)/60))
#clean up
rm(sunriseTimes)

# Read in survey location and habitat data
habitatData <- read_csv("tbl_survey_locations_exported_09_26_2022_pct_suitable_2021.csv")
habitatData$Site_ID <- factor(habitatData$Site_ID)

habitatAndSurveyData <-
habitatData %>%
  select(Site_ID,ln_UTM1083,lt_UTM1083,ln_WGS84,lt_WGS84, pct_suitable_2021) %>%
  right_join(.,surveyData, by = 'Site_ID', keep = F)

habitatAndSurveyData$Site_ID <- factor(habitatAndSurveyData$Site_ID)
habitatAndSurveyData$Observer <- factor(habitatAndSurveyData$Observer)

# Make encounter histories for detected birds across intervals
surveyData$encounterHistory <- paste(if_else(surveyData$Interval_1 == "None", 0,1),
                                     if_else(surveyData$Interval_2 == "None", 0,1),
                                     if_else(surveyData$Interval_3 == "None", 0,1),
                                     if_else(surveyData$Interval_4 == "None", 0,1),
                                     sep = "")

```

```{r Distance sampling frames., echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
## Create a new variable called 'distance', which translates the distance_band information into the midpoint of the distance.
## Then remove all other variables.
  ## THIS EXCLUDES ALL NON-SINGING MALES ##
dists <-
  surveyData %>%
  group_by(Site_ID) %>%
  mutate(distance = ifelse(distanceBand == 1, 12.5,
                           ifelse(distanceBand == 2, 61,
                                  ifelse(distanceBand == 3, 150,
                                         ifelse(distanceBand == 4, 300, NA))))) %>%
  select(Site_ID, distance, Sex, firstDet) %>%
  filter(!is.na(distance), firstDet == "Singing", Sex == "Male")

## Note here that we need the "as.data.frame" argument because 'dists' is a tidyverse tibble, 
## and unmarked doesn't seem to like tibbles. This forces it into a standard R data frame.
yDat <- formatDistData(distData = as.data.frame(dists), distCol = "distance", transectNameCol = "Site_ID", 
                       dist.breaks = c(0,25,100,200,400))

## Create a data frame of site-level covariates.
covs <-
  surveyData %>%
  group_by(Site_ID) %>%
  summarise(site = first(Site_ID),
            observer = first(Observer),
            temp = first(Temp),
            avgNoise = first(Avg_Noise),
            dayOfYear = first (dayOfYear),
            mas = first(mas))


umf <- unmarkedFrameDS(y = as.matrix(yDat), siteCovs = as.data.frame(covs),
                       survey = "point", dist.breaks = c(0,25,100,200,400), unitsIn = "m")
```

## Distance sampling
As expected, most of the detections of singing, male Streaked Horned Larks occured in the middle two distance bands (26-100 m and 101-200 m, respectively).

```{r Plot of detection distances., echo=FALSE, message=FALSE, warning=FALSE}
hist(umf, freq = TRUE, xlab = "Distance (m)", main = "Streaked Horned Lark detections 2022", cex.lab = 0.8, cex.axis = 0.8)
```

Having imported the data and created the necessary data frames, we can fit a series of distance-sampling models. In this example, I've constructed all of the single-factor covariate models using both a half-normal and hazard-rate detection function.

The best-supported model, based on the AICc scores, is the half-normal model that includes day-of-year as a covariate.
```{r Fitting distance models, echo=FALSE, message=FALSE, warning=FALSE}
# Fitting models.
# Half-normal models. 
# Half-normal, null
hnNull <- distsamp(~1~1, umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

#Half-normal, MAS
hnMAS <- distsamp(~mas ~1, umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

# Half-normal, Day of year
hnDay <- distsamp(~dayOfYear ~1, data = umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

# Half-normal, noise
hnNoise <- distsamp(~avgNoise ~1, data = umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

# Half-normal, Temp
hnTemp <- distsamp(~temp ~1, data = umf, keyfun = "halfnorm", output = "density", unitsOut = "kmsq")

# Hazard-rate models
haNull <- distsamp(~1 ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")
haNoise <- distsamp(~avgNoise ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")
haTemp <- distsamp(~temp ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")
haDay <- distsamp(~dayOfYear ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")
haMAS <- distsamp(~mas ~1, data = umf, keyfun = "hazard", output = "density", unitsOut = "kmsq")

fmList <- list("haNull" = haNull, "haDay" = haDay, "haNoise" = haNoise, "haMAS" = haMAS, "haTemp" = haTemp,
               "hnNull" = hnNull, "hnDay" = hnDay, "hnNoise" = hnNoise, "hnMAS" = hnMAS, "hnTemp" = hnTemp)

aictab(cand.set = fmList, second.ord = T, sort = T)
```

Using a parametric bootstrap to examine goodness-of-fit, it seems that this model fits the data adequately. Two of the three tests (Chi-square and Freeman-Tukey) indicated no deviation from expections (both bootstrap P-values > 0.385). However, the test based on the sum of squared errors (SSE) indicates that model fit may not be good. This may be a result of the rather small sample size, so I proceeded even though not all tests agreed that goodness-of-fit was adequate. In the output below, t0 are the observed values of the statistics, and tB are the bootstrapped estimates of the same statistics. In a well-fitting model, the observed and bootstrapped statistics should be similar.

```{r Goodness-of-fit test for best model., echo=FALSE, message=FALSE, warning=FALSE}
# Goodness of fit
fitstats <- function(hnDay) {
  observed <- getY(hnDay@data)
  expected <- fitted(hnDay)
  resids <- residuals(hnDay)
  sse <- sum(resids^2)
  chisq <- sum((observed - expected)^2 / expected)
  freeTuke <- sum((sqrt(observed) - sqrt(expected))^2)
  out <- c(SSE=sse, Chisq=chisq, freemanTukey=freeTuke)
  return(out)
}

pb <- parboot(hnDay, fitstats, nsim=25, report=50)
pb
```

Plotting detectability as a function of day-of-the year, we can see that probability of detecting a lark, given that it is present and available to be detected, declines over the course of the season. Probability of detecting birds further than 100 m from the observer is notably low.
```{r Detectability as a function of day-of-the-year., echo=FALSE, message=FALSE, warning=FALSE}
par(mfrow=c(1, 1))
plot(function(x) gxhn(x, sigma=110.3216), 0, 400, xlab="Distance (m)",
     ylab="Detection probability", cex.lab=1,
     cex.axis=0.7, las=1, col = "green")
plot(function(x) gxhn(x, sigma=46.94988), 0, 400, add = TRUE, col = "blue")
plot(function(x) gxhn(x, sigma=68.06919), 0, 400, add = TRUE, col = "black")
legend('topright', c("First day of surveys", "Last day of surveys", "Median day of surveys"),
       col=c("green", "blue", "black"), lty=1, cex=0.8)
```

The probability (p) of detecting a lark is 0.06 on the median day of surveys, with an effective radius (ER) of 96.2 m.
```{r Estimating detectability from the best model., echo=FALSE, message=FALSE, warning=FALSE}
getPcovs <- function(hnDay) {
  sig <- backTransform(linearComb(hnDay['det'], c(1, median(covs$dayOfYear))))@estimate
  ea <- 2*pi * integrate(grhn, 0, 400, sigma=sig)$value # effective area
  er <- sqrt(ea / pi) # effective radius
  p <- ea / (pi*400^2) #detection probability
  out <- c(p = p , er = er)
  return(out)
}
getPcovs(hnDay)
```

The 95% confidence limits for p and for ER (t_B quantiles 2.5%, 97.5%):
```{r Bootstrap error estimates of P and EDR, echo=FALSE, message=FALSE, warning=FALSE}
getPcovs <- function(hnDay) {
  sig <- backTransform(linearComb(hnDay['det'], c(1, median(covs$dayOfYear))))@estimate
  ea <- 2*pi * integrate(grhn, 0, 400, sigma=sig)$value # effective area
  er <- sqrt(ea / pi) # effective radius
  p <- ea / (pi*400^2) #detection probability
  out <- c(p = p , er = er)
  return(out)
}

parboot(hnDay, getPcovs, nsim = 25, report = 50)
```

The estimated number of larks per square km = 4.33.

```{r Estimate of density from the best distance model, echo=FALSE, message=FALSE, warning=FALSE}
backTransform(hnDay, type = "state")
```

The boostrapped 95% confidence interval around the density estimate (t_B quantiles 2.5%, 97.5%):

```{r Error estimate for density from the best distance model, echo=FALSE, message=FALSE, warning=FALSE}
getD <- function(hnDay) {
  d <- backTransform(hnDay, type = "state")@estimate
  return(d)
}

parboot(hnDay, getD, nsim = 25, report = 50)
```

The estimated density seems significantly higher than current estimates of Streaked Horned Lark population size would predict, probably a result of the very low estimate of detectability.

```{r Joint distance/removal, echo=FALSE, message=FALSE, warning=FALSE}
## Incorporating removal models
encounters <-
  surveyData %>%
  group_by(Site_ID) %>%
  mutate(interval1 = ifelse(firstInterval == 1 & Sex == "Male" & firstDet == "Singing", 1, 0),
         interval2 = ifelse(firstInterval == 2 & Sex == "Male" & firstDet == "Singing", 1, 0),
         interval3 = ifelse(firstInterval == 3 & Sex == "Male" & firstDet == "Singing", 1, 0),
         interval4 = ifelse(firstInterval == 4 & Sex == "Male" & firstDet == "Singing", 1, 0)) %>%
  select(Site_ID, interval1, interval2, interval3, interval4, Sex, firstDet) %>%
  group_by(Site_ID) %>%
  summarise(interval1 = sum(interval1), # this model wants summed # of birds per interval
            interval2 = sum(interval2), # and has to match the distance data
            interval3 = sum(interval3),
            interval4 = sum(interval4)) 

yRemoval <- matrix(nrow = 214, ncol = 4)
rownames(yRemoval) <- encounters$Site_ID
yRemoval <- cbind(encounters[,2:5])
yRemoval[is.na(yRemoval)] <- 0
## Create a new variable called 'distance', which translates the distance_band information into the midpoint of the distance.
## Then remove all other variables.
## THIS INCLUDES ONLY SINGING MALES ##
distsRemoval <-
  surveyData %>%
  group_by(Site_ID) %>%
  mutate(distance = ifelse(distanceBand == 1, 12.5,
                           ifelse(distanceBand == 2, 61,
                                  ifelse(distanceBand == 3, 150,
                                         ifelse(distanceBand == 4, 300, NA))))) %>%
  select(Site_ID, distance, Sex, firstDet) %>%
  filter(!is.na(distance), firstDet == "Singing", Sex == "Male")

## Note here that we need the "as.data.frame" argument because 'dists' is a tidyverse tibble, 
## and unmarked doesn't seem to like tibbles. This forces it into a standard R data frame.
yDatDistance <- formatDistData(distData = as.data.frame(distsRemoval), distCol = "distance", transectNameCol = "Site_ID", 
                       dist.breaks = c(0,25,100,200,400))



umfDR <- unmarkedFrameGDR(yDistance = as.matrix(yDatDistance), yRemoval = as.matrix(yRemoval), numPrimary = 1,
                          siteCovs = covs, dist.breaks = c(0,25,100,200,400), unitsIn = "m")

drDay <- gdistremoval(lambdaformula = ~1, phiformula = ~1, removalformula = ~1,
                       distanceformula = ~dayOfYear, data = umfDR, keyfun = "halfnorm",
                       output = "density", unitsOut = "kmsq", mixture = "ZIP")
```

Using combined distance and removal models, we can estimate detectability:
```{r}
summary(drDay)
getPdistrem <- function(x) {
  sig <- backTransform(linearComb(drDay,c(1,171), type = "dist"))
  ea <- 2*pi * integrate(grhn, 0, 400, sigma=sig@estimate)$value # effective area
  er <- sqrt(ea / pi) # effective radius
  p <- ea / (pi*400^2) #detection probability
  return(p)
}
getPdistrem(drDay)
```
and availability:

```{r Estimating availability from best removal/distance model, echo=FALSE, message=FALSE, warning=FALSE}
backTransform(obj = drDay, type = "rem")
```

and their product:
```{r Estimating detectability, message=FALSE, warning=FALSE}
getPD <- function(x) {
  d <- backTransform(drDay, type = "rem")
  sig <- backTransform(linearComb(drDay,c(1,171), type = "dist"))
  ea <- 2*pi * integrate(grhn, 0, 400, sigma=sig@estimate)$value # effective area
  er <- sqrt(ea / pi) # effective radius
  p <- ea / (pi*400^2) #detection probability
  Pd <- p*(d@estimate)
  return(Pd)
}
getPD(drDay)
```
Not surprisingly, with this exceedingly low probability of detection, the estimate density of larks is quite high:

```{r Density from best removal distance model, echo=FALSE, message=FALSE, warning=FALSE}
backTransform(obj = drDay, type = "lambda")
```

